\chapter{Topological vector space}
\def\O{\mathcal{O}}
\def\V{\mathcal{V}}
\def\N{\mathcal{N}}
\def\T{\mathcal{T}}
\def\F{\mathcal{F}}
 \newcommand{\eqdef}{%
\ensuremath{\mathrel{\stackrel{\mathrm{def}}{=}}}}
\input{sec2.1}
\section{Vector space topologies}
A vector space $V$ which is also a topological space is call a
\emph{topological vector space} if the vector operations
\begin{equation}
  \label{eq:2.2.1}
  V\times V\ni (x,y)\mapsto x+y\in V; \quad K\times V\ni (a,x)\mapsto
  ax\in V
\end{equation}
are continuous. Here and in what follows $K$ denotes the field
$\mathbf{R}$ of real numbers or the field $\mathbf{C}$ of complex
numbers with the usual metric topologies. In particular,
\eqref{eq:2.2.1} implies that for fixed $y$ and fixed $a\in K, a\neq
0$, the maps
\begin{displaymath}
  x\mapsto x+y \quad (\text{translation})\qquad\text{and } x\mapsto ax \quad(\text{dilation})
\end{displaymath}
are homeomorphisms. Thus the open sets are invariant under
translations, so the topology is completely determined by the
neighborhoods of $0$. Since
\begin{displaymath}
  x+x_0+y+y_0=x+y+x_0+y_0,\qquad(a+a_0)(x+x_0)=ax+ax_0+a_0x+a_0x_0,
\end{displaymath}
the continuity of the operations \eqref{eq:2.2.1} reduces to the
following conditions on the neighborhoods of $0$:
\begin{equation*}
  \label{eq:2.2.1m}
  \begin{aligned}
    (x,y) &\mapsto x+y & {\ }&\text{is continuous at }(0,0);\\
    (a,x) &\mapsto ax  & {\ }&\text{is continuous at }(0,0);\\
    a &\mapsto ax_0 & {\ }&\text{is continuous at }0\text{ for every }x_0;\\
    x &\mapsto a_0x & {\ }&\text{is continuous at }0\text{ for every }a_0.
  \end{aligned}\tag*{$(\ref{eq:2.2.1})'$}
\end{equation*}
Explicitly the first three conditions mean that for every neighborhood
$N$ of $0$ in $V$ there shall exist a neighborhood $N_1$ of $0$ and
$\varepsilon>0$ such that
\begin{equation}
  \label{eq:2.2.2}
  \begin{gathered}
    N_1+N_1=\left\{ x+y;\ x,y\in N_1 \right\}\subset N;\quad ax\in
    N\text{ if }|a|\leq\varepsilon\text{ and }x\in N_1;\\
    \text{for every }x\in V\ \exists\varepsilon_x>0\text{ such that }ax\in
    N \quad\text{if }|a|<\varepsilon_x.
  \end{gathered}
\end{equation}
The last condition in \ref{eq:2.2.1m} is a consequence of
\eqref{eq:2.2.2}, for the second part of \eqref{eq:2.2.2} gives
$a_0x\in N$ if $2^kx\in N_1$ for some integer $k>0$ with
$2^{-k}|a_0|<\varepsilon$, and repeated use of the first part of
\eqref{eq:2.2.2} shows that there is a neighborhood $N_2$ of $0$ such
that $2^kN_2\subset N_1$, thus $a_0x\in N$ if $x\in N_2$.

To use the second part of \eqref{eq:2.2.2} we form
\begin{displaymath}
  N_2=\bigcup_{|a|\leq\varepsilon}aN_1\subset N.
\end{displaymath}
This is also a neighborhood of $0$, and we have $aN_2\subset N_2$ if
$|a|\leq1$.
\begin{dfn}
  A set $M$ in a vector space $V$ over $K$ is called \emph{balanced}
  if $ax\in M$ for all $x\in M$ and $a\in K$ with $|a|\leq1$. It is
  called \emph{absorbing} if for every $x\in V$ we have $ax\in M$ when
  $|a|$ is sufficiently small.
\end{dfn}
Every neighborhood of $0$ is absorbing, and we have found that there
is a fundamental system of balanced neighborhoods of $0$. Conversely,
for any system of balanced absorbing sets satisfying the first
condition in \eqref{eq:2.2.2} the finite intersections can be taken as
a fundamental system of neighborhoods of $0$ for a vector space
topology in $V$.

So far we have not required the Hausdorff separation axiom to be
valid. This axiom is equivalent to
\begin{equation}
  \label{eq:2.2.3}
  \text{the intersection of all neighborhoods of }0\text{ is equal to }\left\{ 0 \right\}.
\end{equation}
In fact, \eqref{eq:2.2.3} is obviously a consequence of the Hausdorff
axiom. On the other hand, if \eqref{eq:2.2.3} is valid we can for
every $x\neq0$ find a balanced neighborhood $N$ of $0$ such that
$x\notin N+N$. The neighborhoods $N$ and $\left\{ x \right\}+N$ of $0$
and $x$ are then disjoint, for if $x+y=z;\ y,z\in N$; then $x=z+(-y)\in
N+N$.

It follows from \eqref{eq:2.2.2} that in any topological vector space
$V$ the intersection $V_0$ of all neighborhoods of $0$ is a linear
subspace. If $O$ is an open set in $V$ and if $x\in O, y\in V_0$, it
follows that $x+y\in O$. The open sets in $V$ are therefore unions of
residue classes modulo $V_0$. We obtain a vector space topology in
$V/V_0$ if we take as open sets there the maps of the open sets in $V$
into $V/V_0$; conversely, the open sets in $V$ are the inverse images
of the open sets in $V/V_0$. The definition of $V_0$ and
\eqref{eq:2.2.3} imply that $V/V_0$ satisfies the Hausdorff axiom.

If $W$ is any linear subspace of a topological vector space $V$, we
can also topologize $V/W$ by taking as open sets the images of the
open sets in $V$. It is clear that $V/W$ is a Hausdorff space if and
only if $0$ is closed there, that is, $W$ is a \emph{closed} linear
subspace of $V$.
\begin{thm}\label{thm:2.2.2}
  If $V$ is a finite dimensional Hausdorff topological vector space
  and if $T:K^n\to V$ is a linear bijection, then $T$ is a
  homeomorphism if $K^n$ is given the product topology.
\end{thm}
Thus there is only one Haudorff vector space topology possible in the
finite dimensional case.
\begin{prf}
  Since for some $x_1,\dots,x_n\in V$
  \begin{displaymath}
    Ta=\sum_1^na_jx_j,\quad a=\left( a_1,\dots,a_n \right)\in K^n,
  \end{displaymath}
  it follows from (\ref{eq:2.2.1}) that $T$ is continuous. To prove
  that $T^{-1}$ is continuous we have to show that $T$ maps open sets
  to open sets. This follows if we prove that $TI$ is a neighborhood
  of $0$ when
  \begin{displaymath}
    I=\left\{ a\in K^n;|a_j|<1,j=1,\dots,n \right\}.
  \end{displaymath}
  If $\partial I$ is the boundary of $I$ then
  \begin{displaymath}
    I=\left\{ a\in K^n;wa\notin\partial I \text{ when }w\in K\text{
        and }|w|\leq1 \right\}.
  \end{displaymath}
  Thus
  \begin{displaymath}
    T(I)=\left\{ x\in V;wx\notin T(\partial I)\text{ when }w\in
      K\text{ and }|w|\leq1 \right\}.
  \end{displaymath}
  Since $T(\partial I)$ is a compact set which does not contain $0$
  (Theorem \ref{thm:2.1.3}), it follows, again by Theorem
  \ref{thm:2.1.3}, that there is a neighborhood $N$ of $0$ with $N\cap
  T(\partial I)=\emptyset$. If we take $N$ balanced it follows that
  $N\subset T(I)$.
\end{prf}
\begin{cor}\label{cor:2.2.3}
  Let $W$ be a linear subspace of finite codimension in a topological
  vector space $V$. Then all linear forms vanishing in $W$ are
  continuous if and only if $W$ is closed.
\end{cor}
\begin{prf}
  It follows from Theorem \ref{thm:1.4.2} that $w$ is the intersection
  of the zero sets of linear forms vanishing on $W$, so $W$ is closed
  if these are all continuous. On the other hand, if $W$ is closed
  then $V/W$ is a finite dimensional Hausdorff space. A linear form
  $f$ on $V$ vanishing on $W$ is a composition
  \begin{displaymath}
    V\to V/W\xrightarrow{\ \bar{f}}K
  \end{displaymath}
  where $\bar{f}$ is continuous by Theorem \ref{thm:2.2.2}. Hence $f$
  is also continuous.
\end{prf}
Let $N$ be an open and balanced neighborhood of $0$ in the
topological vector space $V$. Then there is a unique function $p:V\to
\mathbf{R}_{+}$ (the non-negative reals) such that
  \begin{equation}
    \label{eq:2.2.4} p(ax)=|a|p(x)\quad a\in K,x\in V;\qquad N=\left\{
x\in V;p(x)<1 \right\}.
  \end{equation} Moreover, the function $p$ is continuous at $0$. In
fact, if $N$ is defined by $p$ in this way and if $t>0$, then
  \begin{displaymath} p(x)=tp(x/t)\geq t \text{ if } x/t\notin
N;\qquad p(x)<t\text{ if }x/t\in N,
  \end{displaymath} hence
  \begin{equation}
    \label{eq:2.2.5} p(x)=\inf \left\{ t;x/t\in N \right\}.
  \end{equation} Conversely, the function defined by (\ref{eq:2.2.5})
obviously satisfies (\ref{eq:2.2.4}) if $N$ is balanced and
absorbing. If $p(x)<1$ we have $x/t\in N$ for some $t<1$, hence $x\in
N$, and if $p(x)\geq1$ then $x/t\notin N$ if $t<1$. Since $N$ is open
we may conclude that $x\notin N$. Finally $p(x)<\varepsilon$ if
$x\in\varepsilon N$, so $p$ is continuous at $0$.

If $N_1$ is another neighborhood of $0$ of the same kind, and if
$N_1+N_1\subset N$ as in (\ref{eq:2.2.2}),we have $p(x+y)<1$ if
$p_1(x)<1$ and $p_1(y)<1$, where $p_1$ is defined by
(\ref{eq:2.2.5})with $N$ replaced by $N_1$. In view of the homogeneity
it follows that
\begin{displaymath}
  p(x+y)\leq \max\,(p_1(x),\,p_1(y)).
\end{displaymath}
Conversely, if we give a family $\mathcal{P}$ of function from $V$ to
$\mathbf{R}_+$ satisfying (\ref{eq:2.2.4}) such that for every $p\in\mathcal{P}$
\begin{displaymath}
  p(x+y)\leq C\max(p_1(x),p_1(y)),\quad x,y\in V,
\end{displaymath}
for some $p_1\in\mathcal{P}$ and some constant $C$, then the finite
intersections of the sets
\begin{displaymath}
  N_{p,\epsilon}=\left\{ x;p(x)<\epsilon \right\},\qquad\epsilon>0,\quad p\in\mathcal{P},
\end{displaymath}
are a basis of neighborhood of $0$ for a vector space topology in
$V$. The verification is left for the reader.
\begin{exa}
  Let $V$ be the space of $K$ valued continuous functions on
  $[0,1]\subset \mathbf{R}$,and set for some $r>0$
  \begin{displaymath}
    \|f\|_r=\left( \int|f|^r\,dx \right)^{1/r}.
  \end{displaymath}
Then $\|f\|_r$ satisfies (\ref{eq:2.2.4}) and
\begin{displaymath}
  \|f+g\|_r\leq 2^{(r+1)/r}\max(\|f\|_r,\|g\|_r).
\end{displaymath}
Thus the sets $\left\{ f;\|f\|_r<\epsilon \right\}$ form a basis for
neighborhood of $0$ in a vector space topology.
\end{exa}
The general spaces discssed so far are of little use in analysis so we
shall narrow down our discussion to more special classes of spaces
which occur frequently and have useful properties. The first condition
we impose is the existence of a fundamental system of convex
neighborhoods. This is required for the application of Hahn-Banach
theorem proved in Section \ref{sec:1.4}. 
\newcommand{\tvs}{Topological vector spaces}
\newcommand{\lctvs}{Locally convex topological vector spaces}
\newcommand{\sns}{Semi-normed spaces} \newcommand{\ns}{Normed spaces}
\newcommand{\phs}{Pre-Hilbert spaces} \newcommand{\fs}{Fr\'echet
  spaces} \newcommand{\Bs}{Banach spaces} \newcommand{\hs}{Hilbert
  spaces} \newcommand{\comb}{
\begin{tabular}{c}
  \tvs\\$\Downarrow$\\\lctvs
\end{tabular}
}
\begin{center}
  \begin{tabular}{ccccc}
    &\multicolumn{4}{l}{\ \comb}\\
    $\Downarrow$ && \multicolumn{3}{c}{$\Downarrow$}\\
    &&\multicolumn{3}{c}{\sns}\\
    $\Downarrow$ && $\Downarrow$ && $\Downarrow$ \\
    &&\ns && \phs \\
    $\Downarrow$ && $\Downarrow$ && $\Downarrow$ \\
    \fs & $\Longrightarrow$ &\multicolumn{1}{l}{\Bs} &
    $\qquad\Longrightarrow \qquad$ &\multicolumn{1}{l}{\hs}
  \end{tabular}
\end{center}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "alfa"
%%% End: 
