\indentafterchapter
\chapter{Preface}
These lectures might have been called applied linear functional
analysis, for the purpose is not to present functional analysis for
its own sake but rather as a tool for the working analyst. The general
results are therefore illustrated by a large number of examples. They
are on the whole taken from natural contexts. However, functional
analysis alone rarely solves an analytical problemy; its role is to
clarify what is essential in it. It has therefore been necessary to
simplify and modify many of examples so that they can be handled by
elementary arguments. In spite of that the reader may not always have
the necessary background to follow the analytical details, but he
should then rest assured that the examples are not strictly necessary
for the understanding of the main theme.\par
It is not possible to draw a sharp dividing line between analysis and
functional analysis. On the contrary, the vital parts of functional
analysis have always developed from proofs of theorems in
analysis. When some arguments are felt to occur frequently they are
isolated and put in an abstract form, involving rather little
structure so that they can be applied in apparently different
circumstances. This development is still going on, and what is
perceived as central in functional analysis depends to some extent on
what parts of analysis that one is interested in. These lecture notes
are undoubtedly biased towards applications to differential equations
and harmonic analysis, but nevertheless I hope that the material
chosen is of a wide interest. They are a slightly edited version of
lectures given at the University of Lund in 1969 and in 1988.\par
I am grateful to Anders Melin who read the whole manuscript and
suggested a number of improvements. I would also like to thank the
students in the course just finished for their attention to details
which has helped improve the exposition. Some of the exercises added
at the end have been taken over from earlier courses on the same
subject given in Lund or in Corpenhagen.
\vspace*{3em}
\begin{flushright}
  \parbox{5cm}{Lund in February 1989\\\rule{0pt}{5em}Lars
    H\"ormander\hfill}
\end{flushright}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "alfa"
%%% End: 
